---
title: "Contemporaneous and Stationary Dynamic MCMSEM"
author: "MCMSEM developers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contemporaneous and Stationary Dynamic MCMSEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

MCMSEM 0.27.0 exposes the model's time assumption through `kernel`:

```{r}
MCMmodel(ds, kernel = "contemporaneous") # default
MCMmodel(ds, kernel = "dynamic")
```

`kernel = "static"` is a supported, silent alias for `"contemporaneous"`.
Legacy summaries and models without kernel metadata are treated as
contemporaneous. Partial matching is not accepted.

## Contemporaneous Structural MCMSEM

The contemporaneous kernel represents structural equations at one conceptual
occasion, such as (Y = \beta X + \epsilon_Y). Replacing the equation for (X)
changes the value generated for (Y), but the model has no explicit transition
from (t) to (t+1). History, adaptation, feedback, and equilibrium are
absorbed into variables, disturbances, or the path's interpretation. Higher
moments can identify a direction within the assumed structure; they do not add
a temporal order absent from the model.

This kernel is appropriate when a same-occasion intervention is meaningful at
the measurement scale, when one variable plausibly acts before another inside
the measurement window, or when a path is explicitly interpreted as an
integrated or equilibrium response. The existing propagation is unchanged:

\[
M_2 = F_m(I-A)^{-1}S(I-A)^{-T}F_m^T.
\]

## Stationary Dynamic MCMSEM

The dynamic kernel represents repeated, time-homogeneous transitions:

\[
z_t = Bz_{t-1} + \varepsilon_t.
\]

Even when only one stationary cross-section is fitted, the model retains time:

\[
z_t = \varepsilon_t + B\varepsilon_{t-1} + B^2\varepsilon_{t-2} + \cdots.
\]

For order (r), MCMSEM solves

\[
C_r = (I-B^{\otimes r})^{-1}D_r
\]

for orders two, three, and four. Linear solves are used instead of explicit
inverses. Innovation covariance is fixed to (D_2=I). Only diagonal entries of
(D_3) and (D_4) are free, expressing mutually independent innovations with
fixed scale. Stationarity is enforced through \(\rho(B)<1\); autoregressions
default to nonnegative lower bounds while cross-lagged paths remain signed.

The lag interval is substantive. A daily transition and an annual transition
are different parameters. This model is more explicit about dynamics, but also
requires stable repeated transitions, serially independent innovations,
stationarity, and an adequate chosen lag.

## Gaussian residual covariance and fourth moments

The observed cross-section may include an independent Gaussian component,
(y_t=z_t+G), with (G\sim N(0,\Psi_G)). A Cholesky parameterization guarantees
that \(\Psi_G\) is positive semidefinite throughout optimization. It changes
covariance but not third or fourth cumulants. Raw fourth moments use the total
covariance \(\Sigma_y=C_2+\Psi_G\):

\[
M_{4,ijkl}=K_{4,ijkl}+\Sigma_{ij}\Sigma_{kl}
 +\Sigma_{ik}\Sigma_{jl}+\Sigma_{il}\Sigma_{jk}.
\]

## WLS weighting and asymptotic standard errors

Let \(s\) contain the unique raw central moments and let
\(\sigma(\theta)\) contain their dynamic model predictions. The quadratic
criterion is

\[
F(\theta)=(s-\sigma(\theta))^T W(s-\sigma(\theta)).
\]

`MCMdatasummary(..., prep_asymptotic_se = TRUE)` estimates
\(\Omega=\operatorname{Var}(s)\). Its casewise influence functions account for
estimating the sample means in third- and fourth-order central moments. Dynamic
fits support identity weights, inverse marginal variances (`"diagonal"`), or a
regularized \(\Omega^{-1}\) (`"full"`).

For \(\Delta=\partial\sigma(\theta)/\partial\theta^T\), identity and diagonal
weights default to the robust sandwich covariance

\[
(\Delta^T W\Delta)^{-1}\Delta^T W\Omega W\Delta
(\Delta^T W\Delta)^{-1}.
\]

Full WLS defaults to \((\Delta^T\Omega^{-1}\Delta)^{-1}\). The result stores
both robust and model-based matrices; `se_correction` selects which is reported.
SEs for the Cholesky-parameterized \(\Psi_G\) are computed with the delta
method. Because \(\Omega\) here is already the sampling covariance of \(s\), it
already contains the \(1/N\) scale and is not divided by \(N\) again.

Wald SEs are a local, unconstrained approximation. Rank alone is insufficient:
inspect the reported information condition number, and be cautious when a fit
is nearly saturated or a bound/stationarity constraint is active. The bundled
50-replication calibration records good first-order agreement in an
eight-parameter overidentified model and severe conservatism in the weakly
identified 11-parameter bivariate Gaussian-residual model; see
`inst/validation/README.md`.

## Identification

Two variables provide 3 covariance, 4 third-order, and 5 fourth-order unique
moments. The full bivariate dynamic model has 4 transitions, 2 third cumulants,
2 fourth cumulants, and 3 Gaussian covariance parameters: 11 parameters and one
nominal degree of freedom. This count does not prove global identification.
Identification depends on informative and distinct non-Gaussian cumulant
profiles, innovation independence, fixed scale, stationarity, and no omitted
non-Gaussian confounding. Similar or nearly Gaussian innovations can create weak
identification and multiple low-loss states.

The original MCM-SEM framework is described by Zenab Tamimy, Elsje van Bergen,
Matthijs D. van der Zee, Conor V. Dolan, and Michel G. Nivard (2022),
[*Multi Co-Moment Structural Equation Models: Discovering Direction of
Causality in the Presence of
Confounding*](https://doi.org/10.31235/osf.io/ynam2), SocArXiv. The
cumulant-identification framework and discrete Lyapunov formulation underlying
the dynamic kernel build on Cecilie Olesen Recke, Sarah Lumpp, Nataliia
Kushnerchuk, Janike Oldekop, Jiayi Li, Jane Ivy Coons, and Elina Robeva (2026),
[*Identifiability in Graphical Discrete Lyapunov
Models*](https://arxiv.org/abs/2601.21818), arXiv preprint arXiv:2601.21818.

## Workflow

```{r}
ds <- MCMdatasummary(
  dat[, c("X", "Y")],
  scale_data = FALSE,
  use_skewness = TRUE,
  use_kurtosis = TRUE
)
model <- MCMmodel(ds, kernel = "dynamic")
model <- MCMedit(model, "B", c(1, 1), "phi_X")
model <- MCMedit(model, "B", c(1, 2), "Y_lag_to_X")
model <- MCMedit(model, "B", c(2, 1), "X_lag_to_Y")
model <- MCMedit(model, "B", c(2, 2), "phi_Y")
model <- MCMedit(model, "lbound", "phi_X", 0)
model <- MCMedit(model, "lbound", "phi_Y", 0)
fit <- MCMfit(
  model, ds, compute_se = TRUE,
  moment_weighting = "diagonal", n_starts = 5, seed = 2026
)
summary(fit)
MCMdiagnostics(fit, jacobian = TRUE)
```

Use `scale_data = FALSE` when the fixed unit-innovation scale should refer to the
data's stated measurement units. Scaling is allowed, but changes the target
parameterization.

## Scope and future work

This release supports observed-state VAR(1) models with at least two variables,
independent unit-variance innovations, diagonal higher cumulants, optional full
Gaussian residual covariance, moments through order four, WLS weighting, and
asymptotic SEs. Latent measurement models are not yet supported. Dynamic WLS
and SE calculations currently exclude analysis weights and require summaries
prepared by MCMSEM 0.27.0 or later.

A continuous-time analogue is (dz(t)=A z(t)dt+dL(t)), with
(B(\Delta)=\exp(A\Delta)). Future work may add continuous-time cumulants,
VAR(q), combined contemporaneous and lagged paths, nonlinear or regime-switching
dynamics, nonstationarity, latent measurement models, non-Gaussian confounders,
and trajectory-level interventions.
